material and methods

### 2017-09-15 man/ms_1_material_methods.Rmd
update methods

---
title: "material and methods"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../references.bib
csl: ../geb.csl
---

@@

This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight oak patches (2400 Has) identified (:red_circle: FIGURE) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al. 2002, Olalde et al. 2002, Rodríguez-Sánchez et al. 2010) and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change (Hampe and Petit 2005).

@@

:red_circle: La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro). Ver MIGRAME dataset

@@

To characterize the vegetation greeness of *Quercus pyrenaica* we used the Enhanced Vegetation Index (EVI) derived from MOD13Q1 product obtained by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor [@Didan2015]. EVI and NDVI (Normalized Difference Vegetation Index) are the most common greenness vegetation indices. We used EVI instead of NDVI (Normalized Difference Vegetation Index) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002, @Krapivin2015, @Cabello2012].

EVI product consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m. Data were obtained using a Google Earth Engine script (:red_circle: cite gists) for the 2000 - 2016 period. We selected the pixels covering the distribution of Quercus pyrenaica forests in Sierra Nevada (*n* = 928 pixels). The EVI data are geometrically and atmospherically corrected and include information about the quality ass.... 
:red_circle:

@@

@@

### Tree sampling 
Samplig was carried during autumn of 2016. Trees were sampled at two locations located in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Table 1). Both sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..). In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken per tree at 1.3 m using an increment borer. Diameter at breat height (DBH) and total height were recorded for each tree. Increment cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (:red_circle: Rinntech 2003) coupled to a stereomicroscope, with an accuracy of 0.01 mm. Individual ring series were visually and statistically cross-dated with TSAP software (:red_circle: Frank Rinn, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Validation of the croos-dating was done using COFECHA software [@Holmes1983].

## Dendrochronological methods
... dendro
For each focal tree we measured diameter at breast height (DBH) and total height. 
A total of xx trees were sampled. 
...

We built chronologies for each site (two)

>>>>

---
title: "material and methods"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../references.bib
templat
csl: ../geb.csl
---

@@

This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight oak patches (2400 Has) identified (**Figure 1**) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al. 2002, Olalde et al. 2002, Rodríguez-Sánchez et al. 2010) and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change (Hampe and Petit 2005).

@@

@@

To characterize the vegetation greenness of *Q. pyrenaica* we used the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. EVI and NDVI (*Normalized Difference Vegetation Index*) are the most common greenness vegetation indices. We used EVI instead of NDVI because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].

EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.

OJITO---------- reescribir esto

A quality filter was applied to select only valid values of EVI.

First we used information from the Quality Assesment band (QA band) to filter out those values affected by high content of aerosols, clouds, shadows, snow or water;

snow, ice, cloud

and then a quality assessment was carried out to filter the ... [@Reyes2015]

    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds. On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]

EVI data were

The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values. In addition, two types of aerosol loadings typically corrupt EVI—climatology and high aerosols. Use of aerosol climatology

Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2). Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked. Besides, VI usefulness indices between 0 to 11 essentially include all EVI data. Thus, these two conditions serve as additional checks.

The EVI data are geometrically and atmospherically corrected and include information about the quality ass....

@@

OJITO---------- reescribir esto de arriba

@@

### Tree sampling 
Samplig was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1). Two elevations were sampled for the southern site (CA-Low and CA-High). All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..). In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer. For each tree, diameter at breat height (DBH) and total height were recorded. Increment cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.01 mm. Individual ring series were visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Validation of the croos-dating was done using COFECHA software [@Holmes1983].

## Dendrochronological methods
* incluir citas a Frits; Cooks ... 
* site cronologies
* BAI 
* GC Nowacki??
